### 2023-11-21 sections/1-introduction.tex
added new scaling plots and commentary (more to come)
small fixes

{\color{red} We now define two types of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN). 
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \ \setminus \ S_q$, $f(x, q) > \max\{f(y, q): y \in S_q \}$;
that is, for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
On the other hand, $\rho$-nearest neighbors search aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.}

@@

$k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Naive implementations of $k$-NN, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets.
While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not exploit the geometric and topological structure inherent in large datasets.
Approximate search is sufficient for some applications, but the need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$, and small values of $k$ may be necessary when classes are not well-separated.
There is some evidence that distance functions which are non-metrics, such as cosine distance, perform poorly for $k$-NN in biomedical settings~\cite{hu2016distance};
this may suggest that approximate $k$-NN (which may have imperfect precision and recall) could exhibit suboptimal accuracy.

>>>>

We now define two types of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN). 
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \ \setminus \ S_q$, $f(x, q) > \max\{f(y, q): y \in S_q \}$;
that is, for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
On the other hand, $\rho$-nearest neighbors search aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

@@

$k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Na\"{i}ve implementations of $k$-NN, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets.
While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not exploit the geometric and topological structure inherent in large datasets.
Approximate search is sufficient for some applications, but the need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$, and small values of $k$ may be necessary when classes are not well-separated.
There is some evidence that distance functions which are non-metrics, such as cosine distance, perform poorly for $k$-NN in biomedical settings~\cite{hu2016distance};
this may suggest that approximate $k$-NN (which may have imperfect precision and recall) could exhibit suboptimal accuracy.

### 2023-11-21 sections/2-methods.tex
small fixes

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps the query ball, instead of always proceeding with both of its children, we proceed only with those children which also overlap the query ball.

To determine whether both children overlap the query ball, we consider the following geometry. 
Let $c$ denote the center of a cluster, $\ell$ and $r$ its left and right poles (see Section~\ref{subsubsec:methods:cluster-partitioning}), $q$ the query, and $\rho$ the search radius, as shown in Figure [ADD FIGURE].
Without loss of generality, we assume that $\overline{qr} \leq \overline{q\ell}$ (i.e., the query is closer to the right pole than to the left pole).

Now let $m$ be the midpoint of $\ell$ and $r$. Let $d$ denote the distance from $q$ to the perpendicular bisector of $\overline{lr}$, as shown in Figure [ADD FIGURE]. Since we assign a point in the parent cluster to the left child if it is closer to the left pole, if $d > \rho$, (i.e., if $d$ is to the left of the perpendicular bisector of $\overline{\ell r}$), 
the left child does not overlap the query ball. In this case, we proceed only with the right child. If $d \leq \rho$, 
then both child clusters overlap the query ball.

We can do a quick check using values computed at partition time to determine whether $d \leq \rho$.

>>>>

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding with both of its children, we proceed only with those children which could contain points in the query ball.

To determine whether both children can contain points in the query ball, we consider the following geometry. 
Let $\ell$ and $r$ denote the cluster's left and right poles (see Section~\ref{subsubsec:methods:cluster-partitioning}), $q$ denote the query, and $\rho$ denote the search radius, as shown in Figure [ADD FIGURE].
Without loss of generality, we assume that $\overline{qr} \leq \overline{q\ell}$ (i.e., the query is closer to the right pole than to the left pole).

Now let $m$ be the midpoint of $\overline{\ell r}$. Let $d$ denote the distance from $q$ to the perpendicular bisector of $\overline{lr}$, as shown in Figure [ADD FIGURE]. Since we assign a point in the parent cluster to the left child if it is closer to the left pole, if $d > \rho$, 
the left child cannot contain points in the query ball. In this case, we proceed only with the right child. If $d \leq \rho$, 
then both child clusters could contain points in the query ball, and we proceed with both of them.

We can do a quick check using values computed during partitioning (see Section~\ref{subsubsec:methods:cluster-partitioning}) to determine whether $d \leq \rho$. 
We note that $d = ||mq'|| = ||mr||-||q'r||$. Since $m$ is the midpoint of $\overline{\ell r}$, we have that $||mr|| = \tfrac{||\ell r||}{2}$, so it remains to find $||q'r||$. To do this, we use the Law of Cosines with $\triangle \ell rq$. Let $\theta$ denote $\angle lrq$, as shown in Figure [ADD FIGURE]. By the Law of Cosines, we have that 
\begin{equation} \text{cos}(\theta) = \tfrac{||\overline{qr}||^2 + ||\overline{\ell r}||^2 - ||\overline{q\ell}||^2}{2||\overline{\ell r}||}.\end{equation}. 
On the other hand, since $\triangle q'qr$ is a right triangle,

\begin{equation} \text{cos}(\theta) = \tfrac{||\overline{q'r}||}{||\overline{qr}||}.\end{equation}

To check this, it suffices to verify that \begin{equation}(||\overline{q\ell}|| + ||\overline{qr}||)(||\overline{q\ell}||- ||\overline{qr}||) \leq 2||\overline{\ell r}||\rho. \end{equation}
